
% --- Slide 1: Title ---
\begin{frame}[plain]
  \titlepage
\end{frame}
\section{Overview and Big Picture of Lecture 2}

% ---- Learning goals ----
\begin{frame}{Learning goals (what you’ll be able to do)}
\textbf{Goals for today}
\begin{itemize}
\item Pick and configure an optimizer for small control problems (unconstrained \& constrained).
\item Derive KKT conditions and form the SQP/QP subproblems for a nonlinear program.
\item Explain the differences between penalty, augmented Lagrangian, and interior-point methods.
\end{itemize}
\textbf{Why?}\\
In future classes, this will help us map classic control tasks (LQR/MPC/trajectory optimization) to QPs/NLPs and choose a solver strategy.
\end{frame}

% ---- Agenda ----
\begin{frame}{Roadmap for today (2\,hours)}
\begin{enumerate}
\item Big picture and some notation \hfill \textit{(5 min)}
\item Unconstrained optimization: Root-finding, Newton and globalization \hfill \textit{(30 min)}
\item Equality constraints: KKT, Newton vs. Gauss–Newton \hfill \textit{(30 min)}
\item Inequalities \& KKT: complementarity \hfill \textit{(10 min)}
\item Methods: penalty $\rightarrow$ ALM $\rightarrow$ interior-point (PDIP) \hfill \textit{(20 min)}
\item Brief look at SQP for solving hard control problems \hfill \textit{(20 min)} 
\end{enumerate}
\end{frame}



% ---- Big picture ----
\begin{frame}[t]{Big picture: why optimization for control?}
\begin{columns}[T,onlytextwidth]
  \column{0.54\textwidth}
  \small
  \begin{itemize}[<+->]\setlength{\itemsep}{2pt}
    \item Controller synthesis often reduces to solving a sequence of optimization problems.
    \item \textbf{MPC} solves a QP/NLP online at each time step; warm-start and sparsity are critical.
    \item \textbf{Trajectory optimization} (nonlinear robots) uses NLP + collocation; needs robust globalization.
    \item \textbf{Learning-based control} backpropagates through optimizers (differentiable programming).
  \end{itemize}

  \column{0.42\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{%
    \begin{tikzpicture}[
      node distance=10mm and 12mm,
      every node/.style={font=\small},
      shorten >=2pt,
      shorten <=2pt
    ]
      \node[box] (plant) {Dynamics\\$x_{k+1}=f(x_k,u_k)$};
      \node[box, below=of plant] (opt) {Online optimizer\\(QP/NLP)};

      \draw[->] (opt.north) -- node[pos=0.55, right=2pt] {$u_k$} (plant.south);
      \draw[->] (plant.east) -- ++(0.6,0) |- node[pos=0.25, right=2pt] {$x_k$} (opt.east);
    \end{tikzpicture}%
  }
\end{columns}
\end{frame}
